---
title: |
 | Asymmetric Linear Filters for Seasonal Adjustment
 | Applications to the COVID-19
subtitle: "JSM 2021"
author: "Alain Quartier-la-Tente (Lemna, Insee)"
departement: ""
division: |
    | Ongoing research under supervision of:
    | Dominique Ladiray (Independent) and Olivier Darné (Lemna)
    | With the help of Jean Palate (NBB)
logo: "img/logojsm.png"
automaticcontents: false
output:
    beamer_presentation:
        template: template2.tex
        keep_tex: yes
        theme: TorinoTh
        slide_level: 3
        includes:
          in_header: preamble_beamer.tex
themeoptions: "coding=utf8,language=english"
classoption: 'usepdftitle=false,english'
fontsize: 10pt
bibliography: [biblio.bib]
biblio-style: unsrtnat
natbiboptions: [numbers]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE)
library(knitr)
library(kableExtra)
library(rjdfilters)
load(file = "tables.RData")
mae <- readRDS("tables_revisions.RDS")$MAE
```

# Introduction

###  Introduction (1/2)

Ongoing research real-time detection of turning points with linear filters

\faIcon{arrow-circle-right} Under supervision of Dominique Ladiray and Olivier Darné

\faIcon{arrow-circle-right} Help of Jean Palate

\pause

In this presentation:

- First results comparing 4 methods for trend-cycle extraction: 

    - Current X-13ARIMA algorithm (Henderson filter)

    - Local Polynomial filters (@proietti2008)
    
    - Fidelity-Smoothness-Timeliness approach (@ch15HBSA) 
    
    - RKHS filters (@dagumbianconcini2008)

\pause

- \faIcon{r-project} package `rjdfilters` (https://github.com/palatej/rjdfilters, development version: https://github.com/AQLT/rjdfilters)

## Moving averages
###  Introduction (2/2)

A raw time series can be decompose as (additive decomposition):
$$
X_t=\underbrace{TC_t}_{\text{trend-cycle}}+
\underbrace{S_t}_{\text{seasonality}}+
\underbrace{I_t}_{\text{irregular}}
$$
And $TC_t$ generally estimated on a series \highlight{without} seasonality



\pause

\highlight{Moving averages} (or \highlight{linear filters}) are ubiquitous in trend-cycle extraction and seasonal adjustment (e.g.: X-13-ARIMA):
$$
M_\theta(X_t)=\sum_{k=-p}^{+f}\theta_kX_{t+k}
$$
\pause 
\faArrowCircleRight{} In general, \highlight{symmetric} moving averages ($p=f$ et $\theta_{-i}=\theta_i$) 

\pause
\faArrowCircleRight{} For \highlightbf{real-time estimates}, we must rely on \highlight{asymmetric} filters ($p>f$): revisions and delay in turning points detection (\highlight{phase-shift}): this is the case of the COVID-19

\bigskip \pause
\faArrowCircleRight{} Comparison of 3 methods that could be included in X-13-ARIMA


# Description of the methods

### Contents
\tableofcontents[currentsection, hideothersubsections]

## Current approache

### X-13-ARIMA

1. Series extend over 1 year by ARIMA model

2. Trend-Cycle component extracted using symmetric \highlightbf{Henderson} moving average

\pause
\faArrowCircleRight{} Forecasts linear combinations of past values: equivalent to the use of asymmetric filters

\bigskip\pause
\faArrowCircleRight{} X-11: iteratively decomposes $X_T$ in $TC_t$, $S_t$ and $I_t$ with automatic outlier correction

\bigskip\pause
\faArrowCircleRight{} Comparison of 3 alternatives modern approaches that can reproduce Henderson filter


## Local Polynomials

### Local polynomials: \faIcon{r-project} `rjdfilters::lp_filter()`

Assumption: $y_t=\mu_t+\varepsilon_t$ with $\varepsilon_t\overset{i.i.d}{\sim}\mathcal N(0,\sigma^2)$

$\mu_t$ locally approximated by polynomial of degree $d$:
$$
\forall j\in\left\llbracket -h,h\right\rrbracket : y_{t+j}=m_{t+j}+\varepsilon_{t+j},\quad m_{t+j}=\sum_{i=0}^{d}\beta_{i}j^{i}
$$

\pause
Estimation using WLS using *kernels*: $\hat{\beta}=(X'KX)^{1}X'Ky$ and
$$
\hat{m}_{t}=\hat\beta_0=w'y=\sum_{j=-h}^{h}w_{j}y_{t-j}
\text{ \faArrowCircleRight{} equivalent to symmetric moving average}
$$
\faArrowCircleRight{} Henderson filter using a specific kernel and $d=3$.

### Asymmetric filters: \faIcon{r-project} `rjdfilters::lp_filter()`

Several solutions:

1. Same method with less data (DAF) $\iff$ Minimize revisions under same polynomial constraints (reproduce cubic trend)

\faArrowCircleRight{} **no bias** but **lots of variance**  
 

\pause

2. Minimization of revisions filter under polynomial constraints:

    1. *Linear-Constant* (LC): $y_t$ linear and $v$ reproduce constant trends (\highlight{Musgrave} filter)
    
    2. *Quadratic-Linear* (QL): $y_t$ quadratic and $v$ reproduce linear trends
    
    3. *Cubic-Quadratic* (CQ): $y_t$ cubic and $v$ reproduce quadratic trends
    
    \faArrowCircleRight{} Asymmetric filters $v$ depends on a ratio linked to "IC-Ratio"
    
\pause
\bcsmbh simple models with easy interpretation

\bcsmmh Timeliness not controlled \faArrowCircleRight{} method extended in `rjdfilters::lp_filter()`



## Linear Filters and Reproducing Kernel Hilbert Space (RKHS)

### RKHS filters: \faIcon{r-project} `rjdfilters::rkhs_filter()`

- RKHS theory used to approximate Henderson filter

- With $K_p$ the **kernel function**, the symmetric filter:
$$
\forall j\in\left\llbracket -h,h\right\rrbracket: w_{j}=\frac{K_p(j/b)}{\sum_{i=-h}^{^h}K_p(i/b)}
$$

\onslide<3->{\faArrowCircleRight{} with $b=h+1$ and a specific $K_p$ you have the Henderson filter}

\pause

- For asymmetric filters:
$$
\forall j\in\left\llbracket -h,q\right\rrbracket: w_{a,j}=\frac{K_p(j/b)}{\sum_{i=-h}^{^q}K_p(i/b)}
$$

\pause\pause\faArrowCircleRight{} $b$ chosen by optimization, e.g. minimizing revisions linked to phase-shift:
$$b_{q,\varphi}=\underset{b_q}{\min}
\int_{0}^{2\pi/12}
\rho_s(\lambda)\rho_\theta(\lambda)\sin^{2}\left(\frac{\varphi_\theta(\omega)}{2}\right)\ud \omega$$

### Filtres asymétriques

::::{.columns}
:::{.column width=65%}

\bcsmmh several local extremum

\footnotesize
```{r, out.height="50%", fig.align="center"}
fun <- rkhs_optimization_fun(horizon = 6, 
            leads = 5, degree = 3, 
            asymmetricCriterion = "Timeliness")
plot(fun, 5.6, 12, xlab = "b", 
     ylab = "Timeliness", main = "6X5 filter")
rkhs_optimal_bw()
```
:::
:::{.column width=30%}

\bigskip

\bcsmbh Generalizable to create filters that could be applied to irregular frequency series

:::
::::


## Minimization under constraints: FST approach

### FST approach: \faIcon{r-project} `rjdfilters::fst_filter()`

Minimization of a weighted sum of 3 criteria under polynomial constraints:

$$
\begin{cases}
\underset{\theta}{\min} & J(\theta)=
\alpha F_g(\theta)+\beta S_g(\theta)+\gamma T_g(\theta)\\
s.c. & C\theta=a
\end{cases}
$$
$F_g$: Fidelity (variance reduction ratio), $S_g$: Smoothness (Henderson criterion), $T_g$ timeliness (phase-shift)

\pause
::: {.summary data-latex=""}
- \bcsmbh Unique solution
  
- \bcsmbh Asymmetric filters independent of data and symmetric filter
  
- \bcsmmh Non-normalized weights
:::


# Comparison of the methods

### Contents
\tableofcontents[currentsection, hideothersubsections]

## Methodology
### Methodology

2 404 calendar adjusted series (`sts_inpr_m`, industrial production indices of EU):

1. Seasonal adjustment with X-13ARIMA (`RJDemetra::x13`) for each date to extract: linearized component, length of trend and seasonal filters, decomposition mode and I-C ratio

\pause

2. Seasonal adjustment using *fixed* linearized component (est. overall period) and same length/decomposition/I-C ratio with X-11 using custom trend-cycle filters (`rjdfilters::x11()`).

\pause

3. For each estimate, downturns and upturns:

    - upturn: $y_{t-3}\geq\cdots\geq y_{t-1}<y_t\leq y_{t+1}$
    
    - downturn: $y_{t-3}\leq\cdots\leq y_{t-1}>y_t\geq y_{t+1}$
    
\pause

Time-delay: time to detect the correct turning point without further revisions
    
Methods compared: RKHS minimizing phase-shift and local polynomial filters
    
## Time delay
### Time delay to detect turning points in 2020

For series for which the optimal trend-cycle symmetric filter is of length 13 (900 series)

\footnotesize

```{r covid-quantile, echo = FALSE, eval = FALSE}
kbl(table_covid,
    booktabs = TRUE,
    align = c("c",rep("c", ncol(table_covid))))  %>% 
    kable_styling(latex_options = c("striped"))
```

\alt<2->{\newcolumntype{C}{>{\columncolor{processblue!30}}c}}{\newcolumntype{C}{c}}

\alt<2>{\newcolumntype{D}{>{\columncolor{processblue!30}}c}}{\newcolumntype{D}{c}}

\alt<3>{\newcolumntype{E}{>{\columncolor{processblue!30}}c}}{\newcolumntype{E}{c}}
\alt<4>{\newcolumntype{F}{>{\columncolor{processblue!30}}c}}{\newcolumntype{F}{c}}
\begin{table}
\centering
\begin{tabular}[t]{lDFDEEE}
\toprule
  & \textbf<2>{X-13-ARIMA} & \textbf<4>{RKHS} & \textbf<2>{LC} & \textbf<3>{QL} & \textbf<3>{CQ} & \textbf<3>{DAF}\\
\midrule
{Min} & {2.0} & {2.00} & {2.00} & {2.00} & {2.00} & {2.00}\\
D1 & 2.0 & 5.00 & 3.00 & 2.00 & 2.00 & 2.00\\
{D2} & {3.0} & {5.00} & {3.00} & {3.00} & {2.00} & {2.00}\\
D3 & 3.0 & 5.00 & 4.00 & 3.00 & 3.00 & 3.00\\
{D4} & {3.0} & {5.00} & {4.00} & {3.00} & {5.00} & {3.00}\\
Median & 4.0 & 5.00 & 4.00 & 4.00 & 6.00 & 5.00\\
{D6} & {4.0} & {7.00} & {4.00} & {5.00} & {6.00} & {6.00}\\
D7 & 4.0 & 7.00 & 5.00 & 7.00 & 7.00 & 7.00\\
{D8} & {5.0} & {7.00} & {5.00} & {7.00} & {7.00} & {7.00}\\
D9 & 9.0 & 9.00 & 8.00 & 9.00 & 9.00 & 9.00\\
{Max} & {14.0} & {14.00} & {14.00} & {14.00} & {14.00} & {14.00}\\
Mean & 4.4 & 6.29 & 4.69 & 4.97 & 5.32 & 5.09\\
\bottomrule
\end{tabular}
\end{table}


## Revision
### MAE in 2020

For series for which the optimal trend-cycle symmetric filter is of length 13

$R_t$: relative revision error between first and last estimates.

Distribution of $\frac{MAE(R_t)}{MAE(R_t^{X-13})}$

\footnotesize

```{r, echo = FALSE, eval = FALSE}
kbl(mae,
    booktabs = TRUE,
    align = c("c",rep("c", ncol(mae))))  %>% 
    kable_styling(latex_options = c("striped"))
```


\alt<2->{\newcolumntype{C}{>{\columncolor{processblue!30}}c}}{\newcolumntype{C}{c}}

\alt<2>{\newcolumntype{D}{>{\columncolor{processblue!30}}c}}{\newcolumntype{D}{c}}

\alt<3>{\newcolumntype{E}{>{\columncolor{processblue!30}}c}}{\newcolumntype{E}{c}}
\alt<4>{\newcolumntype{F}{>{\columncolor{processblue!30}}c}}{\newcolumntype{F}{c}}
\begin{table}
\centering
\begin{tabular}[t]{lDEFFF}
\toprule
  & \textbf<2>{RKHS} & \textbf<3>{LC} & \textbf<4>{QL} & \textbf<4>{CQ} & \textbf<4>{DAF}\\
\midrule
{Min} & {0.4} & {0.5} & {0.5} & {0.5} & {0.5}\\
D1 & 0.9 & 0.9 & 1.1 & 1.1 & 1.1\\
{D2} & {1.0} & {1.0} & {1.3} & {1.4} & {1.3}\\
D3 & 1.1 & 1.1 & 1.4 & 1.5 & 1.5\\
{D4} & {1.1} & {1.2} & {1.6} & {1.7} & {1.6}\\
Median & 1.2 & 1.3 & 1.7 & 1.8 & 1.8\\
{D6} & {1.3} & {1.4} & {1.9} & {2.0} & {1.9}\\
D7 & 1.4 & 1.5 & 2.0 & 2.3 & 2.2\\
{D8} & {1.5} & {1.6} & {2.3} & {2.6} & {2.4}\\
D9 & 1.7 & 1.8 & 2.6 & 3.1 & 2.9\\
{Max} & {4.0} & {4.6} & {6.5} & {5.8} & {5.9}\\
Mean & 1.3 & 1.3 & 1.8 & 2.0 & 1.9\\
\bottomrule
\end{tabular}
\end{table}

## Some examples



### IPI in the manufacture of fabricated metal products, except machinery and equipment (C25) in Sweden (turning point in February 2020)

\centering
\includegraphics[height = 0.5\paperheight]{img/C25SE_x13}


###

\centering
\includegraphics[height = 0.5\paperheight]{img/C25SE_rkhs} 
\includegraphics[height = 0.5\paperheight]{img/C25SE_lc}

###

\centering
\includegraphics[height = 0.5\paperheight]{img/C25SE_ql} 
\includegraphics[height = 0.5\paperheight]{img/C25SE_cq}

###

\centering
\includegraphics[height = 0.5\paperheight]{img/C25SE_daf}

### IPI in the manufacture of cement, lime and plaster (C235) in Germany (turning point in February 2020)


\centering
\includegraphics[height = 0.5\paperheight]{img/C235DE_x13}

###

\centering
\includegraphics[height = 0.5\paperheight]{img/C235DE_rkhs} 
\includegraphics[height = 0.5\paperheight]{img/C235DE_lc}

###

\centering
\includegraphics[height = 0.5\paperheight]{img/C235DE_ql} 
\includegraphics[height = 0.5\paperheight]{img/C235DE_cq}

###

\centering
\includegraphics[height = 0.5\paperheight]{img/C235DE_daf} 


# Conclusion

## Conclusion
### Conclusion

- To build asymmetric filters, we can focus on the ones that reproduces at most linear trend (excluding QL, CQ and DAF filters)

\bigskip \pause

- During the COVID-19, the current X-13-ARIMA algorithm seems to produce on average satisfying results


\bigskip \pause

- In some cases, we could prefer others trend-cycle filters \faIcon{arrow-circle-right} `rjdfilters` can help


### What next\bcquestion

- More investigations to understand why and when a method performs better

\pause
- Study on other datasets

\pause

Other methods:

- FST can lead to filters that performs better in terms of Fidelity, Smoothness, Timeliness than:

    - RKHS filters with same polynomial constraints

    - LC filters with same polynomial constraints
    
    \faIcon{arrow-circle-right} Study of those moving averages?
    

\pause
- Direct Filter Approach (@trilemmaWMR2019), Cascade Linear Filter (@cascadeFilter), etc.

\pause
- Impact of outliers? Study of robust methods?



<!-- ``` {=latex} -->
<!-- \end{frame} -->
<!-- \appendix -->
<!-- \begin{frame}{Bibliography} -->
<!-- ``` -->

### Thanks for your attention


\begin{columns}
\begin{column}{0.6\textwidth} 
\faIcon{r-project} package: \href{https://github.com/palatej/rjdfilters}{\faGithub{} palatej/rjdfilters}
\end{column}
\begin{column}{0.4\textwidth}
About me: \href{https://github.com/AQLT}{\faGithub{} AQLT}  
\end{column}
\end{columns}

**Bibliography**:
\footnotesize


