---
title: "RJDemetra: an R interface to JDemetra+"
subtitle: "useR!2019"
author: "Alain Quartier-la-Tente"
division: "Insee, Seasonal Adjustment Centre of Excellence"
departement: "alain.quartier-la-tente@insee.fr"
logo: "img/SACElogo.jpg"
automaticcontents: true
output:
    beamer_presentation:
        template: template.tex
        keep_tex: yes
        theme: TorinoTh
        slide_level: 3
header-includes:
- \usepackage{wrapfig}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage[table]{xcolor}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{animate}
- \usepackage{fontawesome5}
themeoptions: "coding=utf8,language=english"
classoption: 'usepdftitle=false,french'
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                        fig.path = "img/markdown-")
library(AQLTools)
library(RJDemetra)
options(enable_print_style = FALSE)
```


# Introduction to seasonal adjustment

### Introduction to seasonal adjustment (1/3)

```{r, echo = FALSE, fig.height = 6,fig.cap="Monthly CRAN downloads of ggplot2"}
dataGraph <- readRDS("ggplot2.RDS")
ggplot(data = dataGraph, aes(x = date, y = value)) +
  geom_line(size=0.70) + theme_bw() + 
  labs(x = NULL, y = "Monthly downloads (/1000)",
       title = "ggplot2")
```


### Introduction to seasonal adjustment (2/3)

\begin{table}
\begin{tabular}{cc}
\toprule
Weekday & Average CRAN downloads \\
\midrule
Monday & 14628\\
Tuesday & 15941\\
Wednesday & 15706\\
Thursday & 15206\\
Friday & 12948\\
Saturday & 7038\\
Sunday & 7104\\
\bottomrule
\end{tabular}
\caption{Average of CRAN downloads of ggplot2 per weekday since 2015}
\end{table}

### Introduction to seasonal adjustment (3/3)

Purpose of seasonal adjustment:

- Time comparison (outlook, short-term evolution...)  
- Spatial comparison

\bigskip
\pause
Two leading methods:

- TRAMO/SEATS+ (Bank of Spain)  
- X-12ARIMA/X-13ARIMA-SEATS (US-Census Bureau).

\bigskip
\pause

$\rightarrow$ proceed in two steps:

1. Pre-adjusting the series of deterministics effects with a RegARIMA model

2. Decomposition: to extract seasonal component

### What's JDemetra+\bcquestion 

\includegraphics[width=2cm]{img/jdemetra+.png} TRAMO/SEATS+ and X-13ARIMA-SEATS are implemented in JDemetra+ (JD+)

\bigskip
\large\faThumbsUp{} \normalsize Software [officially recommended](https://ec.europa.eu/eurostat/cros/system/files/Jdemetra_%20release.pdf) by Eurostat and the ECB for seasonal and calendar adjustment of official statistics

\bigskip
$\rightarrow$ RJDemetra is an \large\faRProject{} \normalsize interface to JDemetra+ based on the \large\faJava{} \normalsize libraries of JD+


# RJDemetra

## Current status
### Current status

- RegARIMA, TRAMO-SEATS and X-13-ARIMA:  
    + pre-defined and user-defined specifications: outliers detection, ARIMA detection, userdefined regressors, transformation function...  
    + S3 classes with plot, summary, print methods
    
\medskip

- Manipulate JD+ workspaces:  
    + Import JD+ workspace to get input raw series or SA model
    + Export R models created via RJDemetra
    
\medskip

- Include a dataset: industrial production indices in manufacturing in the European Union

### Object structure

A `SA` object is a `list()` of 5 elements:

\includegraphics{img/sa_obj_struct.png}

### Create your first model

Like in JD+ users can defined their own specification or use a pre-defined one:

\footnotesize
```{r}
library(RJDemetra)
ipi_fr <- ipi_c_eu[, "FR"]
ts_mod <- tramoseats(ipi_fr, spec = "RSAfull")
x13_usr_spec <- x13_spec(spec = c("RSA5c"),
                         usrdef.outliersEnabled = TRUE,
                         usrdef.outliersType = c("LS", "AO"),
                         usrdef.outliersDate = c("2008-10-01",
                                                 "2002-01-01"),
                         usrdef.outliersCoef = c(36, 14),
                         transform.function = "None")
x13_mod <- x13(ipi_fr, x13_usr_spec, userdefined = "diagnostics.ic-ratio")
```

Use `user_defined_variables()` to get the names of the user-defined variables

## RegARIMA examples

### RegARIMA examples (1/2)
\footnotesize

```{r}
summary(x13_mod$regarima)
```

### RegARIMA examples (2/2)

```{r, echo = FALSE}
def.par <- par(no.readonly = TRUE)
par(mai = c(0,0.82,0,0.42))
```

```{r, fig.height = 6 }
layout(matrix(1:6, 3, 2));plot(x13_mod$regarima, ask = FALSE)
```

## Seasonal adjustment examples

### Seasonal adjustment examples (1/7): decomposition

\footnotesize

```{r}
x13_mod$decomposition
```

### Seasonal adjustment examples (2/7): decomposition

\footnotesize

```{r}
ts_mod$decomposition
```


### Seasonal adjustment examples (3/7)

```{r, fig.height=6}
plot(x13_mod$decomposition)
```

### Seasonal adjustment examples (4/7)

\footnotesize

```{r}
x13_mod$final
```

### Seasonal adjustment examples (5/7)

```{r,fig.height=6}
plot(x13_mod$final, first_date = 2012, type_chart = "sa-trend")
```

### Seasonal adjustment examples (6/7)

```{r,fig.height=6}
plot(x13_mod$final, last_date = 2000, type_chart = "cal-seas-irr")
```

### Seasonal adjustment examples (7/7)

\footnotesize
```{r,fig.height=6}
x13_mod$diagnostics
```

## Export a JD+ workspace
### Export a workspace

\footnotesize
```{r, echo= TRUE, message = FALSE}
wk <- new_workspace()
new_multiprocessing(wk, name = "MP-1")
add_sa_item(wk, multiprocessing = "MP-1",
            sa_obj = x13_mod, name =  "SA with X13 model 1 ")
add_sa_item(wk, multiprocessing =  "MP-1",
            sa_obj = ts_mod, name = "SA with TramoSeats model 1")
save_workspace(wk, "workspace.xml")
```

\includegraphics{img/workspace.png}

## Import a JD+ workspace

### Import a workspace
\footnotesize

```{r}
wk <- load_workspace("workspace.xml")
compute(wk) # Important to get the Sa model
models <- get_model(wk, progress_bar = FALSE) # get all models
# Or to get one specific model:
mp <- get_object(wk, 1)
count(mp)
sa2 <- get_object(mp, 2)
get_name(sa2)
mod <- get_model(sa2, wk)
```

## Reduce time computation

### Manipulate \faJava{} objects (1/2)

\footnotesize
Default functions can be time consuming (computation of outputs)... Especially if you only need one specific parameter

$\rightarrow$ "Manipulate" java models: `jx13`, `jtramoseats`, `jregarima`, `jregarima_x13`, `jregarima_tramoseats` and `get_jmodel`

```{r}
jx13_mod <- jx13(ipi_fr, x13_usr_spec)
# To get the available outputs:
tail(get_dictionary(jx13_mod), 2)
# To get an indicator:
get_indicators(jx13_mod, "diagnostics.ic-ratio")
# To get the previous R output
x13_mod <- jSA2R(jx13_mod)
```

$\rightarrow$ The output can be customize by every user/institute  

# Around RJDemetra and JDemetra+

## Around RJDemetra

### Examples of current use of RJDemetra

- ggdemetra: ggplot2 extension for 'RJDemetra' 

\faGithub{} https://github.com/AQLT/rjdqa

- rjdqa: package to help quality assessment (dashboard and quality report matrix)

\faGithub{} https://github.com/AQLT/rjdqa

- persephone: enable easy processing during production of SA series (interactive plots, dashboards...)

\faGithub{} https://github.com/statistikat/persephone

- rjdmarkown: nice rmarkdown outputs for RJDemetra

\faGithub{} https://github.com/AQLT/rjdmarkdown

- Carry out studies on SA: Ladiray D., Quartier-la-Tente A., "(In)Stability of Reg-ARIMA Models for Seasonal Adjustment"

### rjdqa
\footnotesize
```{r,fig.height=6.5, echo = TRUE}
plot(rjdqa::sa_dashboard(x13_mod))
```

### ggdemetra


## Around JDemetra+
### Around JDemetra+
- State space framework of JD+:  
\faGithub{} https://github.com/nbbrd/rjdssf

\medskip

- Benchmarking and temporal disaggregation:  
\faGithub{} https://github.com/palatej/rjdbench

\medskip

- R interface to the JWSACruncher (console tool to refresh the models of a JD+ workspace:  
\faGithub{} https://github.com/AQLT/rjwsacruncher


# Installation and future developments
## How to install the package?

### How to install the package?

The package is available on \large\faGithub\normalsize: https://github.com/jdemetra/rjdemetra

It has also it's own website: https://jdemetra.github.io/rjdemetra/

```{r, eval = FALSE}
# Cran release
install.packages("RJDemetra")

# Development version
devtools::install_github("jdemetra/rjdemetra")
```

\bcinfo To install it you need Java8:  in case you don't, install a portable version of Java8 and set the `JAVA_HOME` path.

See the installation manual: https://github.com/jdemetra/rjdemetra/wiki/Installation-manual

## Why use RJDemetra?

### Why use RJDemetra \bcquestion

- Methods used are recommended by Eurostat

- Performance and integration in production with JDemetra+

- Lots of \large\faRProject{} \normalsize developments around `RJDemetra`

- `RJDemetra` evolves with JDemetra+: will integrate new developments on SA methods

## Future developments

### What's next? \bcpanchant

- documentation: article for the Journal of Statistical Software + cheat sheet

- shiny app to change the specification

With JD+ 3.0.0 (by the end of 2020):

- Function to "refresh" the model

- Compatibility with all frequencies (JD+ daily, weekly, etc.)

### Thank you for your attention...

... And don't forget your stickers!

\bigskip

\begin{columns}
\begin{column}{0.33\textwidth}
\begin{center}
\includegraphics[width=\textwidth]{img/rjdemetra_logo.png}
\end{center}
\end{column}
\begin{column}{0.33\textwidth}
\begin{center}
\includegraphics[width=3cm]{img/jdemetra+.png}
\end{center}
\end{column}
\begin{column}{0.33\textwidth} 
\begin{center}
\includegraphics[width=3cm]{img/ggdemetra_logo.png}
\end{center}
\end{column}
\end{columns}

\bigskip
\bigskip

\begin{columns}
\begin{column}{0.5\textwidth}
Alain Quartier-la-Tente


\href{mailto:alain.quartier-la-tente@insee.fr}{\faEnvelope{} alain.quartier-la-tente@insee.fr} 
\end{column}
\begin{column}{0.5\textwidth} 
\href{https://github.com/jdemetra/rjdemetra}{\faGithub{} jdemetra/rjdemetra}  

\href{https://twitter.com/JDemetraPlus}{\faTwitter{} @JdemetraPlus}

\end{column}
\end{columns}


