---
title: "rjd3modelling and documentation"
subtitle: "TSACE, October 26, 2021"
author: "Alain Quartier-la-Tente"
division: "Insee)"
departement: "alain.quartier-la-tente@insee.fr"
logo: "img/SACElogo"
output:
    beamer_presentation:
        template: template.tex
        keep_tex: yes
        theme: TorinoTh
        slide_level: 3
header-includes:
- \usepackage{wrapfig}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage[table]{xcolor}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{animate}
- \usepackage{fontawesome5}
themeoptions: "coding=utf8,language=english"
classoption: 'usepdftitle=false,french'
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                        fig.path = "img/markdown-")
library(AQLTools)
```


# rjd3modelling

### What is available in `rjd3modelling`?

Package available on GitHub:
```{r, eval=FALSE}
remotes::install_github("palatej/rjd3toolkit")
remotes::install_github("palatej/rjd3modelling")
```

. . .

- create trading-days variables with a **user-defined calendar**: easter related days (`calendar.easter`), fixed days (`calendar.fixedday`) and from specific holidays `calendar.holiday` \faArrowCircleRight{} see `?calendar.new` for a complete example.

. . .

- create **common regressors**: stock trading days (`stock.td`), leap year regressors (`lp.variable`),
easter regressors (`easter.variable`),
outliers (`ao.variable`, `ls.variable`, `tc.variable`, `so.variable`), ramp (`ramp.variable`),  intervention variables (`intervention.variable`), periodic dummies (`periodic.dummies`) and contrast (`periodic.contrasts`), trigonometric variables (`trigonometric.variables`)

. . .

\bcquestion How do you get started with these tools?

# Documentation
## Static document

### Vignette/Word/PDF documentation

Usual documentation, already available for JDemetra+

\animategraphics[loop, autoplay, width=\linewidth]{2}{img/gif/docx/docx_}{1}{8}

### HTML/PDF tutorials with `unilur` {.allowframebreaks}

Use `unilur` (github.com/koncina/unilur) to create tutorials/practicals or examination papers with `rmarkdown`

```
---
output:
  unilur::tutorial_html: default # without solutions
  unilur::tutorial_html_solution: default # with solutions
  unilur::tutorial_pdf: default
  unilur::tutorial_pdf_solutionn: default
---
```

````md
```{r, solution = TRUE}`r ''`
frenchCalendar <- calendar.new()
```
````

\animategraphics[loop, autoplay, width=\linewidth]{2}{img/gif/unilur/unilur_}{1}{23}

## Dynamic tutorial

### Create interactive tutorials with `learnr` {.allowframebreaks}

Tutorials includes in a R package (`remotes::install_github("AQLT/rjd3tutorials")`)

````md
```{r regressors, exercise = TRUE}`r ''`
frenchCalendar <- calendar.new()
```

```{r regressors-hint}`r ''`
# define Saturday and Sunday as contrast
groups <- c(1, 1, 1, 1, 1, 0, 0)
```

```{r regressors-solution}`r ''`

```

````

\animategraphics[loop, autoplay, width=\linewidth]{2}{img/gif/learnr/learnr_}{1}{44}

### Thank you for your attention


\begin{columns}
\begin{column}{0.4\textwidth}
\faIcon{r-project} packages:
\begin{itemize}
\item \href{https://github.com/palatej/rjd3toolkit}{\faGithub{} palatej/rjd3toolkit}

\item \href{https://github.com/palatej/rjd3modelling}{\faGithub{} palatej/rjd3modelling}
\end{itemize}
\end{column}
\begin{column}{0.6\textwidth}
Tutorials:
\begin{itemize}
\item \href{https://github.com/AQLT/rjd3tutorials}{\faGithub{} AQLT/rjd3tutorials}

\item example with unilur: \href{https://aqlt-formation-rte.netlify.app/TP/Enonces/R-2-CJO_solution.html}{https://aqlt-formation-rte.netlify.app/TP/Enonces/R-2-CJO\_solution.html}
\end{itemize}
\end{column}
\end{columns}



\vfill
Contact:
\href{mailto:alain.quartier-la-tente@insee.fr}{\faEnvelope{} alain.quartier-la-tente [at] insee.fr}



