---
title: "RJDemetra: an R interface to JDemetra+"
subtitle: "SAEG & SACE, 11 March 2019"
author: "Alain Quartier-la-Tente and Anna Michalek"
division: "Insee, Seasonal Adjustment Centre of Excellence (AQLT) and European Central Bank (AM)"
departement: "alain.quartier-la-tente@insee.fr"
logo: "img/SACElogo.jpg"
output:
    beamer_presentation:
        template: template.tex
        keep_tex: yes
        theme: TorinoTh
        slide_level: 3
header-includes:
- \usepackage{wrapfig}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage[table]{xcolor}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{animate}
- \usepackage{fontawesome5}
themeoptions: "coding=utf8,language=english"
classoption: 'usepdftitle=false,french'
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                        fig.path = "img/markdown-")
library(AQLTools)
library(RJDemetra)
options(enable_print_style = FALSE)
```


# RJDemetra

## Purpose and current status

### Purpose of the RJDemetra package


RJDemetra is an \large\faRProject{} \normalsize interface to JDemetra+ based on the \large\faJava{} \normalsize libraries of JDemetra+


* Complete R package for Tramo-Seats and X13  
* Users: "pure R" package
    + Part of R routines, automatization
        - Batch processing
        - E.g.: direct vs indirect aggregates adjustment, dashboards
    + Usage of other R functions and packages
    
* JD+ functionality 
    + Modeling and seasonal adjustment 
    + Full specification
    
* Advanced graphical presentation: JD+ 

### Current status

- RegARIMA, TRAMO-SEATS and X-13-ARIMA:  
    + pre-defined and user-defined specifications  
    + S3 classes with plot, summary, print methods
    
\medskip
   
- Manipulate JD+ workspaces:  
    + Import JD+ workspace to get input raw series or SA model
    + Export R models created via RJDemetra
    
\medskip

- Include a dataset: industrial production indices in manufacturing in the European Union


## RegARIMA examples

### RegARIMA examples (1/3)

```{r}
library(RJDemetra)
ipi_fr <- ipi_c_eu[,"FR"]
regarima_model <- regarima_def_x13(ipi_fr, spec = "RG4c")
regarima_model
```

### RegARIMA examples (2/3)
\footnotesize

```{r}
summary(regarima_model)
```

### RegARIMA examples (3/3)

```{r, echo = FALSE}
def.par <- par(no.readonly = TRUE)
par(mai = c(0,0.82,0,0.42))
```

```{r, fig.height = 6 }
layout(matrix(1:6, 3, 2));plot(regarima_model, ask = FALSE)
```


## Seasonal adjustment examples
### Seasonal adjustment examples (1/8)

A `SA` object is a `list()` of 5 elements:

\includegraphics{img/sa_obj_struct.png}

### Seasonal adjustment examples (2/8)

Like in JD+ users can defined their own specification or use a pre-defined one:

\footnotesize
```{r}
x13_usr_spec <- x13_spec_def(spec = c("RSA5c"),
                             usrdef.outliersEnabled = TRUE,
                             usrdef.outliersType = c("LS", "AO"),
                             usrdef.outliersDate = c("2008-10-01",
                                                     "2002-01-01"),
                             usrdef.outliersCoef = c(36, 14),
                             transform.function = "None")
x13_mod <- x13(ipi_fr, x13_usr_spec)
ts_mod <- tramoseats_def(ipi_fr, spec = "RSAfull")
```


### Seasonal adjustment examples (3/8): decomposition

\footnotesize

```{r}
x13_mod$decomposition
```

### Seasonal adjustment examples (4/8): decomposition

\footnotesize

```{r}
ts_mod$decomposition
```


### Seasonal adjustment examples (5/8)

```{r, fig.height=6}
plot(x13_mod$decomposition)
```

### Seasonal adjustment examples (6/8)

\footnotesize

```{r}
x13_mod$final
```

### Seasonal adjustment examples (7/8)

```{r,fig.height=6}
plot(x13_mod$final, first_date = 2012, type_chart = "sa-trend")
```


### Seasonal adjustment examples (8/8)

\footnotesize
```{r,fig.height=6}
x13_mod$diagnostics
```



## Manipulate workspaces


### Export a workspace

\footnotesize
```{r, echo= TRUE, message = FALSE}
wk <- new_workspace()
new_multiprocessing(wk, name = "MP-1")
add_sa_item(wk, multiprocessing = "MP-1",
            sa_obj = x13_mod, name =  "SA with X13 model 1 ")
add_sa_item(wk, multiprocessing =  "MP-1",
            sa_obj = ts_mod, name = "SA with TramoSeats model 1")
save_workspace(wk, "workspace.xml")
```

\includegraphics{img/workspace.png}

### Import a workspace (1/3)

\footnotesize

```{r}
wk <- load_workspace("workspace.xml")
get_ts(wk)
```

### Import a workspace (2/3)

\animategraphics[loop, autoplay, width=\linewidth]{2.5}{img/gif/import_model/}{1}{114}



### Import a workspace (3/3)
\footnotesize

```{r}
compute(wk) # Important to get the Sa model
models <- get_model(wk) # A progress bar is printed by default
# To extract only one model
mp <- get_object(wk, 1)
count(mp)
sa2 <- get_object(mp,2)
get_name(sa2)
mod <- get_model(wk, sa2)
```


## How to install the package?

### How to install the package?

The package is available on \large\faGithub\normalsize: https://github.com/jdemetra/rjdemetra

It has also it's own website: https://jdemetra.github.io/rjdemetra/

It package can be installed from CRAN:
```{r, eval = FALSE}
install.packages("RJDemetra")
```

Or from github (development version):

```{r, eval = FALSE}
devtools::install_github("jdemetra/rjdemetra")
```

\bcinfo To install it you need Java8:  in case you don't, install a portable version of Java8 and set the `JAVA_HOME` path.

## Future developments

### What's next? \bcpanchant (1/2)

Documentation:

- Vignette/article for the Journal of Statistical Software

- Guide to install the package with portable version of Java (when you don't have administrator rights)

- Cheat sheet

### What's next? \bcpanchant (2/2)

Package:

- Get only the Java object of a SA (to reduce computation/customize the output)

- Possibility to used user-defined calendar regressors (currently: only user-defined regressors)

- Function to "refresh" the model (JD+ 3.0.0)

# How to use RJDemetra to improve production of SA series?

## Around RJDemetra

### Examples of current use of RJDemetra

- rjdqa (experimental, no documentation): package to help quality assessment (dashboard and quality report matrix)

\faGithub{} https://github.com/AQLT/rjdqa

- persephone (experimental): enable easy processing during production of SA series (interactive plots, dashboards...)

\faGithub{} https://github.com/statistikat/persephone

- Non explore topics: direct vs indirect adjustment (persephone), analyse of revisions, etc.


- Carry out studies on SA: Ladiray D., Quartier-la-Tente A., "(In)Stability of Reg-ARIMA Models for Seasonal Adjustment" $\rightarrow$ STS05 in room MANS

### rjdqa
\footnotesize
```{r,fig.height=6.5, echo = TRUE}
plot(rjdqa::sa_dashboard(x13_mod))
```

### persephone

\animategraphics[loop, autoplay, width=0.9\linewidth]{3}{img/gif/persephone/}{1}{167}

## Around JDemetra+
### Around JDemetra+
- Interface around JD+ 3.0:  
\faGithub{} https://github.com/nbbrd/jd3-rtests

\medskip

- State space framework of JD+:  
\faGithub{} https://github.com/palatej/rjdssf

\medskip

- Benchmarking and temporal disaggregation with JD+ 3.0:  
\faGithub{} https://github.com/palatej/rjdbench

\medskip

- R interface to the JWSACruncher (no documentation yet):  
\faGithub{} https://github.com/AQLT/rjwsacruncher


### Thank you for your attention

\vspace{-0.2cm}

\begin{columns}
\begin{column}{0.4\textwidth}
\begin{center}
\includegraphics[width=4cm]{img/rjdemetra_logo.png}
\end{center}
\end{column}
\begin{column}{0.5\textwidth} 
\href{https://github.com/jdemetra/rjdemetra}{\faGithub{} jdemetra/rjdemetra}  

\href{https://twitter.com/JDemetraPlus}{\faTwitter{} @JdemetraPlus}

Other works and packages around JD+:  
\href{https://github.com/nbbrd}{\faGithub{} nbbrd}  
\end{column}
\end{columns}



\vfill
Contact:  
Alain Quartier-la-Tente  
\href{mailto:alain.quartier-la-tente@insee.fr}{\faEnvelope{} alain.quartier-la-tente@insee.fr}  



