---
title: "RJDemetra: an R interface to JDemetra+"
subtitle: "Hackathon on RJDemetra: 1 & 2 July"
author: "Alain Quartier-la-Tente and Anna Michalek"
division: "Insee, Seasonal Adjustment Centre of Excellence (AQLT) and European Central Bank (AM)"
departement: "alain.quartier-la-tente@insee.fr"
logo: "img/SACElogo.jpg"
output:
    beamer_presentation:
        template: template.tex
        keep_tex: yes
        theme: TorinoTh
        slide_level: 3
header-includes:
- \usepackage{wrapfig}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage[table]{xcolor}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{animate}
- \usepackage{fontawesome5}
themeoptions: "coding=utf8,language=english"
classoption: 'usepdftitle=false,french'
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                        fig.path = "img/markdown-")
library(AQLTools)
library(RJDemetra)
options(enable_print_style = FALSE)
```


# RJDemetra

## Purpose and current status

### Purpose of the RJDemetra package

(slides: https://github.com/AQLT/slides folder "2019 - 07 - Hackathon Frankfurt")


RJDemetra is an \large\faRProject{} \normalsize interface to JDemetra+ based on the \large\faJava{} \normalsize libraries of JDemetra+


* Complete R package for Tramo-Seats and X13  
* Users: "pure R" package
    + Part of R routines, automatization
        - Batch processing
        - E.g.: direct vs indirect aggregates adjustment, dashboards
    + Usage of other R functions and packages
    
* JD+ functionality 
    + Modeling and seasonal adjustment 
    + Full specification
    
* Advanced graphical presentation: JD+ 

### Current status

- RegARIMA, TRAMO-SEATS and X-13-ARIMA:  
    + pre-defined and user-defined specifications  
    + S3 classes with plot, summary, print methods
    
$\rightarrow$ Programing-choice: not manipulate the Java object and use S3 classes (more simple)
    
\pause
\medskip
   
- Manipulate JD+ workspaces:  
    + Import JD+ workspace to get input raw series or SA model
    + Export R models created via RJDemetra
    
\medskip

- Include a dataset: industrial production indices in manufacturing in the European Union


## RegARIMA examples

### RegARIMA examples (1/3)

```{r}
library(RJDemetra)
ipi_fr <- ipi_c_eu[,"FR"]
regarima_model <- regarima_x13(ipi_fr, spec = "RG4c")
regarima_model
```

### RegARIMA examples (2/3)
\footnotesize

```{r}
summary(regarima_model)
```

### RegARIMA examples (3/3)

```{r, echo = FALSE}
def.par <- par(no.readonly = TRUE)
par(mai = c(0,0.82,0,0.42))
```

```{r, fig.height = 6 }
layout(matrix(1:6, 3, 2));plot(regarima_model, ask = FALSE)
```


## Seasonal adjustment examples
### Seasonal adjustment examples (1/9)

A `SA` object is a `list()` of 5 elements:

\includegraphics{img/sa_obj_struct.png}

### Seasonal adjustment examples (2/9)

Like in JD+ users can defined their own specification or use a pre-defined one:

\footnotesize
```{r}
x13_usr_spec <- x13_spec(spec = c("RSA5c"),
                         usrdef.outliersEnabled = TRUE,
                         usrdef.outliersType = c("LS", "AO"),
                         usrdef.outliersDate = c("2008-10-01",
                                                 "2002-01-01"),
                         usrdef.outliersCoef = c(36, 14),
                         transform.function = "None")
x13_mod <- x13(ipi_fr, x13_usr_spec)
ts_mod <- tramoseats(ipi_fr, spec = "RSAfull")
```


### Seasonal adjustment examples (3/9): decomposition

\footnotesize

```{r}
x13_mod$decomposition
```

### Seasonal adjustment examples (4/9): decomposition

\footnotesize

```{r}
ts_mod$decomposition
```


### Seasonal adjustment examples (5/9)

```{r, fig.height=6}
plot(x13_mod$decomposition)
```

### Seasonal adjustment examples (6/9)

\footnotesize

```{r}
x13_mod$final
```

### Seasonal adjustment examples (7/9)

```{r,fig.height=6}
plot(x13_mod$final, first_date = 2012, type_chart = "sa-trend")
```

### Seasonal adjustment examples (8/9)

```{r,fig.height=6}
plot(x13_mod$final, last_date = 2000, type_chart = "cal-seas-irr")
```

### Seasonal adjustment examples (9/9)

\footnotesize
```{r,fig.height=6}
x13_mod$diagnostics
```

## User-defined variable

To get user-defined variables you can use the parameter `userdefined` in `x13()` and `tramoseats()`. To get the available one:
```{r,fig.height=6}
head(user_defined_variables("X13-ARIMA"))
tail(user_defined_variables("TRAMO-SEATS"))
x13_mod <- x13(ipi_fr,
               userdefined = c("diagnostics.ic-ratio", "diagnostics.msr(1)"))
x13_mod$user_defined
x13_mod$user_defined$`diagnostics.msr(1)`
```


# Manipulate workspaces

## Export a workspace
### Export a workspace

\footnotesize
```{r, echo= TRUE, message = FALSE}
wk <- new_workspace()
new_multiprocessing(wk, name = "MP-1")
add_sa_item(wk, multiprocessing = "MP-1",
            sa_obj = x13_mod, name =  "SA with X13 model 1 ")
add_sa_item(wk, multiprocessing =  "MP-1",
            sa_obj = ts_mod, name = "SA with TramoSeats model 1")
save_workspace(wk, "workspace.xml")
```

\includegraphics{img/workspace.png}

## Import a workspace

### Import a workspace (1/4)

\footnotesize

```{r}
wk <- load_workspace("workspace.xml")
get_ts(wk)
```

### Import a workspace (2/4)

\animategraphics[loop, autoplay, width=\linewidth]{2.5}{img/gif/import_model/}{1}{114}

### Import a workspace (3/4)
\footnotesize

```{r}
compute(wk) # Important to get the Sa model
models <- get_model(wk) # A progress bar is printed by default
# To extract only one model
mp <- get_object(wk, 1)
count(mp)
sa2 <- get_object(mp,2)
get_name(sa2)
mod <- get_model(wk, sa2)
```

### Import a workspace (4/4)

Known issues:

- Import may be incomplete: models with Ramp effects, IV, benchmarking... partially imported: \bcsmbh correct results but \bcsmmh changing the spec will erase them

$\rightarrow$ \bcquestion Someone has a \faRProject{} function to create IV?

- `"X11"` spec not implemented: will not be imported $\rightarrow$ needed?

- Error when there is "metadata" (comments, etc.): https://github.com/jdemetra/rjdemetra/issues/53

# Reduce time computation

## Manipulate Java objects

### Manipulate \faJava{} objects (1/2)

\footnotesize
Default functions can be time consuming (computation of outputs)... Especially if you only need one specific parameter

\medskip

$\rightarrow$ "Manipulate" java models: `jx13`, `jtramoseats`, `jregarima`, `jregarima_x13`, `jregarima_tramoseats` and `get_jmodel`

\medskip

```{r}
jx13_mod <- jx13(ipi_fr, x13_usr_spec)
# To get the available outputs:
tail(get_dictionary(jx13_mod))
```


### Manipulate \faJava{} objects (2/2)

\footnotesize
```{r}
# To get an indicator:
get_indicators(jx13_mod, "diagnostics.td-res-all", "diagnostics.ic-ratio")
# To get the previous R output
x13_mod <- jSA2R(jx13_mod)
```

$\rightarrow$ The output can be customize by every user/institute  

\medskip

\bctakecare No error returned by `jx13()` with "wrong" SA (preliminary check...) and `get_indicators()` returns `NULL` objects

\medskip

\bcinfo `add_sa_item` not compatible with `jSA` objects

## Benchmarking

### Bencharking with X-13 on French IPI

\footnotesize
`r R.version["version.string"]`, `r R.version["platform"]`, Windows 7 x64 (build 7601) Service Pack 1

```{r,echo=FALSE,message=FALSE,warning=FALSE, fig.height = 6}
library(microbenchmark)
library(ggplot2)
autoplot(readRDS("microbenchmark.RDS")) + theme(text = element_text(size=20))
```


# Installation and future developments
## How to install the package?

### How to install the package?

The package is available on \large\faGithub\normalsize: https://github.com/jdemetra/rjdemetra

It has also it's own website: https://jdemetra.github.io/rjdemetra/

```{r, eval = FALSE}
# Cran release
install.packages("RJDemetra")

# Development version
devtools::install_github("jdemetra/rjdemetra")
```

\bcinfo To install it you need Java8:  in case you don't, install a portable version of Java8 and set the `JAVA_HOME` path.

See the installation manual: https://github.com/jdemetra/rjdemetra/wiki/Installation-manual

## Future developments

### What's next? \bcpanchant

- Vignette/article for the Journal of Statistical Software. Work in progress: https://github.com/AQLT/vignette_rjdemetra

- Cheat sheet

- Benchmarking?

- Function to "refresh" the model (JD+ 3.0.0)

# Around RJDemetra

### Examples of current use of RJDemetra

- rjdqa: package to help quality assessment (dashboard and quality report matrix)

\faGithub{} https://github.com/AQLT/rjdqa

- persephone: enable easy processing during production of SA series (interactive plots, dashboards...)

\faGithub{} https://github.com/statistikat/persephone

- ggdemetra: ggplot2 extension for 'RJDemetra' 

\faGithub{} https://github.com/AQLT/rjdqa

- Carry out studies on SA: Ladiray D., Quartier-la-Tente A., "(In)Stability of Reg-ARIMA Models for Seasonal Adjustment"

- Non explore topics: we will see after the hackathon

### Ideas for the hackathon

- Export/import specifications as a data.frame

- RJDemetran and chain-linking (KIX)

- Revision Analysis

- Tool for direct/indirect adjustment

- Shiny app for interactively changing specification 

- Automatic report of a model/workspace: PDF, markdown (html, docx...)

- Additional features importing/exporting workpaces: handle error when there are comments, get metadata (and set?), import of intervention variables/Ramp effect


### Thank you for your attention

\vspace{-0.2cm}

\begin{columns}
\begin{column}{0.4\textwidth}
\begin{center}
\includegraphics[width=4cm]{img/rjdemetra_logo.png}
\end{center}
\end{column}
\begin{column}{0.5\textwidth} 
\href{https://github.com/jdemetra/rjdemetra}{\faGithub{} jdemetra/rjdemetra}  

\href{https://twitter.com/JDemetraPlus}{\faTwitter{} @JdemetraPlus}

Other works and packages around JD+:  
\href{https://github.com/nbbrd}{\faGithub{} nbbrd}  
\end{column}
\end{columns}



\vfill
Contact:  
Alain Quartier-la-Tente  
\href{mailto:alain.quartier-la-tente@insee.fr}{\faEnvelope{} alain.quartier-la-tente@insee.fr}  



