---
title: |
 | R and JDemetra+ 3.0: 
 | A new toolbox around seasonal adjustment and time series analysis
subtitle: "uRos2022"
author: "Alain Quartier-la-Tente"
departement: "06/11/2022"
division: | 
    | Insee
    | Scientific Session: Time series and longitudinal data analysis
logo: "img/logobeamer.png"
automaticcontents: true
automaticcontentstext: "Contents"
output:
    beamer_presentation:
        template: template_Beamer.tex
        keep_tex: yes
        theme: TorinoTh
        slide_level: 3
        includes:
          in_header: preamble_beamer.tex
themeoptions: "coding=utf8,language=english"
classoption: 'usepdftitle=false,french' #handout
fontsize: 10pt
bibliography: [biblio.bib]
biblio-style: unsrtnat
natbiboptions: [numbers]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE, 
                      fig.align = 'center',
                      fig.path = "img/slides-")
library(knitr)
library(kableExtra)
library(microbenchmark)
library(ggplot2)
library(patchwork)
library(rjd3toolkit)
library(rjd3modelling)
library(rjd3sa)
library(rjd3arima)
library(rjd3x13)
library(rjd3tramoseats)
library(rjdemetra3)
library(ggdemetra3)
rjd3toolkit::reload_dictionaries()

# library(rjd3sts)
# library(rjd3highfreq)

# library(rjd3bench)
fig.ext <- "pdf"
is_html <- knitr::is_html_output()
is_latex <- knitr::is_latex_output()
fig.ext = "pdf"
if(is_html){
    fa_arrow_circle_right <- '<i class="fas fa-arrow-circle-right"></i>'
    fa_r_project <- '<i class="fab fa-r-project"></i>'
}else {
    if(is_latex){
        fa_arrow_circle_right <- "\\faIcon{arrow-circle-right}"
        fa_r_project <- "\\faIcon{r-project}"
        fa_java <- "\\faIcon{java}"
    }else {
        fa_arrow_circle_right <- "->"
        fa_r_project <- 'R'
    }
}
if (!dir.exists("benchmark"))
  dir.create("benchmark")

f_bm_airline <- "benchmark/airline.RDS"
f_bm_arima <- "benchmark/arima_complicated.RDS"
f_bm_seasonal <- "benchmark/seasonal.RDS"
if (!file.exists(f_bm_airline)) {
  mairline = microbenchmark::microbenchmark("stats::arima" = {
   m = arima(log(AirPassengers), order=c(0,1,1), seasonal = list(order=c(0,1,1), period=12))
},
"rjd3modelling::sarima.estimate" = {
  rjd3modelling::sarima.estimate(log(AirPassengers), order=c(0,1,1), seasonal = list(order=c(0,1,1), period=12))
})
  # autoplot(mairline)
  saveRDS(mairline, f_bm_airline)
}
if (!file.exists(f_bm_arima)) {
  mcomplicates = microbenchmark::microbenchmark("stats::arima" = {
  arima(log(AirPassengers), order=c(3,1,1), seasonal = list(order=c(1,1,1), period=12))
},
"rjd3modelling::sarima.estimate" = {
  rjd3modelling::sarima.estimate(log(AirPassengers), order=c(3,1,1), seasonal = list(order=c(1,1,1), period=12))
})
  # autoplot(mcomplicates)
  saveRDS(mcomplicates, f_bm_arima)
}

if (!file.exists(f_bm_seasonal)) {
  mseasonal <- microbenchmark::microbenchmark(
    "seasonal::seas" = seasonal::seas(AirPassengers, x11 = ""),
    "RJDemetra::x13" = RJDemetra::x13(AirPassengers),
    "RJDemetra::jx13" = RJDemetra::jx13(AirPassengers),
    "rjd3x13::x13" = rjd3x13::x13(AirPassengers,"rsa5c"),
    "rjd3x13::fast.x13" = rjd3x13::fast.x13(AirPassengers,"rsa5c"),
    "rjd3x13::jx13" = rjd3x13::jx13(AirPassengers,"rsa5c")
  )
  # autoplot(mseasonal)
  saveRDS(mseasonal,
			f_bm_seasonal)
}

options(width = 60)
def.par <- par(no.readonly = TRUE)
```

# Introduction

### JDemetra+ \bcquestion

- JDemetra+ is an open source software (build on `r fa_java`) officially recommended by Eurostat for seasonal adjustment (SA)

. . .

- Implements the two leading SA methods X-13ARIMA and TRAMO-SEATS with a nice graphical interface



### Introduction (1)

-   In March 2019, `RJDemetra` was published on CRAN:

    -   only `r fa_r_project` package that enables to use TRAMO-SEATS

    -   faster than existing `r fa_r_project` packages on seasonal adjustment

    -   enables to interact with JDemetra+ "workspaces" used in production
    
. . .

- Other `r fa_r_project` packages around SA

  - `stats::stl()` and `forecast::mstl()` for (M)STL: not recommended because they cannot perform calendar adjustment  
  - `seasonal` and `x12`: interface to X-13-ARIMA-SEATS US Census Bureau binaries

. . .

-   With the development of JDemetra+ 3.0, more than 13 `r fa_r_project` packages are being developped! Not only on seasonal adjustment!

. . .

- They require Java `r fa_java` $\geq$ 17 (see for example installation manual of `RJDemetra`: <https://github.com/jdemetra/rjdemetra/wiki/Installation-manual>) 


### Introduction (2)

They are all available in GitHub, currently:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("palatej/rjd3toolkit")
remotes::install_github("palatej/rjd3modelling")
remotes::install_github("palatej/rjd3sa")
remotes::install_github("palatej/rjd3arima")
remotes::install_github("palatej/rjd3x13")
remotes::install_github("palatej/rjd3tramoseats")
remotes::install_github("palatej/rjdemetra3")
remotes::install_github("palatej/rjdfilters")
remotes::install_github("palatej/rjd3sts")
remotes::install_github("palatej/rjd3highfreq")
remotes::install_github("palatej/rjd3stl")
remotes::install_github("palatej/rjd3bench")
remotes::install_github("AQLT/ggdemetra3")
```

### Introduction (3)

![](img/diag.pdf)

And it's just the begining! (might change in the future)

# Utility packages

## rjd3toolkit

### rjd3toolkit

Contains several utility functions used in other `rjd` packages and several functions to perform tests:

- Normality tests: Bowman-Shenton (`bowmanshenton()`), Doornik-Hansen (`doornikhansen()`), Jarque-Bera (`jarquebera()`, with more parameters than `tseries::jarque.bera.test()`)

- \bclampe Runs tests (randomness of data):  mean or the median (`testofruns()`) or up and down runs test (`testofupdownruns()`)

- autocorrelation functions (usual, inverse, partial)

- `aggregate()` to aggregate a time serie to a higher frequency


## rjd3modelling

### rjd3modelling

\footnotesize

- \bclampe create user-defined calendar and trading-days regressors: 
  `calendar.new()` (create a new calendar), `calendar.holiday()` (add a specific holiday, e.g. christmas),
  `calendar.easter()` (easter related day) and `calendar.fixedday()`)

. . .

- \bclampe create outliers regressors (AO, LS, TC, SO, Ramp, intervention variables), calendar related regressors (stock, leap year, periodic dummies and contrasts, trigonometric variables) -> to be added quadratic ramps

. . .

- \bclampe  Range-mean regression test (to choose log transformation, `rangemean.tstat()`), Canova-Hansen (`td.ch()`) and trading-days f-test (`td.f()`)

. . .

- manipulation of ARIMA models (generation, sum, decomposition, estimation)

. . .

- functions to stationarise your series `do.stationary()`, `differences()`, `differencing.fast()`

. . .

- specification functions for `rjd3x13` and `rjd3tramoseats`


### Example of a specific calendar  {.allowframebreaks}

```{r}
library(rjd3modelling)
fr_cal <- calendar.new()
calendar.holiday(fr_cal, "NEWYEAR")
calendar.holiday(fr_cal, "EASTERMONDAY")
calendar.holiday(fr_cal, "MAYDAY")
calendar.fixedday(fr_cal, month = 5, day = 8,
                  start = "1953-03-20")
# calendar.holiday(fr_cal, "WHITMONDAY") # Equivalent to:
calendar.easter(fr_cal, offset = 61)

calendar.fixedday(fr_cal, month = 7, day = 14)
# calendar.holiday(fr_cal, "ASSUMPTION")
calendar.easter(fr_cal, offset = 61)
calendar.holiday(fr_cal, "ALLSAINTSDAY")
calendar.holiday(fr_cal, "ARMISTICE")
calendar.holiday(fr_cal, "CHRISTMAS")
```

\footnotesize
Use `holidays()` to get the days of the holidays and `htd()` to get the trading days regressors

```{r}
holidays(fr_cal, "2020-12-24", 10, single = TRUE)
s <- ts(0, start = 2020, end = c(2020, 11), frequency = 12)
# Trading-days regressors (each day has a different effect, sunday as contrasts)
td_reg <- htd(fr_cal, s = s, groups = c(1, 2, 3, 4, 5, 6, 0))
# Working-days regressors (Monday = ... = Friday; Saturday = Sunday = contrasts)
wd_reg <- htd(fr_cal, s = s, groups = c(1, 1, 1, 1, 1, 0, 0))
# Monday = ... = Friday; Saturday; Sunday = contrasts
wd_reg <- htd(fr_cal, s = s, groups = c(1, 1, 1, 1, 1, 2, 0))
wd_reg
```


### Example of outliers {.allowframebreaks} 

\footnotesize

```{r outplot}
s <- ts(0, start = 2000, end = 2005, frequency = 12)
ao <- ao.variable(s = s, date = "2001-03-01")
ls <- ls.variable(s = s, date = "2001-01-01")
tc <- tc.variable(s = s, date = "2001-01-01", rate = 0.7)
so <- so.variable(s = s, date = "2003-05-01")
ramp <- ramp.variable(s = s, range = c("2001-01-01", "2001-12-01"))
plot(ts.union(ao, ls, tc, so, ramp), plot.type = "single",
     col = c("red","lightgreen","orange","blue","black"))
```


### Benchmark of ARIMA estimations

More than 20 time faster in median!

\footnotesize

```{r performance-arima, echo = FALSE, out.height="90%", message=FALSE}
(autoplot(readRDS(f_bm_airline), log = TRUE) + ggtitle("ARIMA(0,1,1)(0,1,1)")) /
(autoplot(readRDS(f_bm_arima), log = TRUE) + ggtitle("ARIMA(3,1,1)(1,1,1)"))
```


## rjd3sa
### rjd3sa (1)

Seasonality tests:

- Canova-Hansen (`seasonality.canovahansen()`)

- \bclampe X-12 combined test (`seasonality.combined()`)

- F-test on seasonal dummies (`seasonality.f()`)

- Friedman Seasonality Test (`seasonality.friedman()`)

- Kruskall-Wallis Seasonality Test (`seasonality.kruskalwallis()`)

- \bclampe Periodogram Seasonality Test (`seasonality.periodogram()`)

- QS Seasonality Test (`seasonality.qs()`)


### rjd3sa (2)
\bcattention Always correct the trend and remove the mean before seasonality tests:
\footnotesize
```{r}
library(rjd3sa)
y <- diff(rjd3toolkit::ABS$X0.2.09.10.M, 1); y <- y - mean(y)
# Or:
y <- rjd3modelling::differences(rjd3toolkit::ABS$X0.2.09.10.M)
seasonality.f(y)
seasonality.friedman(y)
seasonality.kruskalwallis(y)
seasonality.combined(y)
```


# Seasonal adjustment packages


## rjd3arima
### rjd3arima

`rjd3arima` is devoted to formatting the output of Arima related results 

## rjd3x13 and rjd3tramoseats

### Common functions


Common functions (defined in `rjd3modelling`) to set the specification of the preprocessing:

`set_arima()`, `set_automodel()`, `set_basic()`, `set_easter()`, `set_estimate()`, `set_outlier()`, `set_tradingdays()`, `set_transform()`, `add_outlier()` and `remove_outlier()`, `add_ramp()`, `remove_ramp()`, `add_usrdefvar()`

### rjd3x13

Main functions:

- Specification: created with `spec_x11_default()`, `spec_x13_default()`, `spec_regarima_default()` and customized with `rjd3modelling` functions + `set_x11()`

- Apply SA model with `x11()`, `x13()`, `fast.x13()`

- ARIMA modelling with `regarima()`, `fast.regarima()`

- \bclampe Refresh policies: `regarima.refresh()` and `x13.refresh()`


### Performance

In median: `RJDemetra` more 3 time faster than `seasonal` and `rjdemetra3` more than 12 time faster than `seasonal`!

```{r performance, echo = FALSE, out.height="90%", message=FALSE}
m <- readRDS(f_bm_seasonal)
autoplot(m)+theme(axis.text=element_text(size=16),
        axis.title=element_text(size=16))
```


### Exemple {.allowframebreaks}
\scriptsize

```{r x13, fig.dim=c(6,3)}
library(rjd3modelling);library(rjd3x13)
y <- rjd3toolkit::ABS$X0.2.09.10.M
spec <- spec_x13_default("rsa5c") |> set_easter(type = "unused") |> 
  set_outlier(outliers.type = c("AO", "LS")) |> 
  set_tradingdays(test = "None") |> set_x11(henderson.filter = 13) |> 
  add_outlier(type = "TC", date = "2000-06-01",
              name = "My TC in 2000-06") 
m = rjd3x13::x13(y, spec)
m$result$preprocessing
# Also summary function
# summary(m)
plot(m)
```


## rjd3tramoseats
### rjd3tramoseats

Main functions:

- Specification: created with `spec_tramoseats_default()`, `spec_tramo_default()` and customized with `rjd3arima` functions + `set_seats()`

- Apply model with `tramoseats()`, `fast.tramoseats()`, `tramo()`, `fast.tramo()`

- \bclampe Refresh policies: `tramo.refresh()` and `tramoseats.refresh()`

Example:


```{r tramoseats, out.height="70%"}
spec <- spec_tramoseats_default("rsafull") |> 
  set_easter(type = "IncludeEasterMonday") |> 
  set_tradingdays(test = "Separate_T") |> 
  set_seats(algorithm = "KalmanSmoother")
m <- rjd3tramoseats::tramoseats(y, spec)
```



## rjdemetra3

### rjdemetra3

Functions to manipulate JDemetra+ workspaces:

- Still in construction: you can load an existing workspace but not create a new one (use `jws.load()` for example)

- Will contain all the functionalities of `rjdworkspace` (more manipulation of workspaces)


## rjd3highfreq and rjd3stl

### rjd3highfreq and rjd3stl

Seasonal adjustment of high frequency data:

- \bclampe fractional and multi airline decomposition 

- \bclampe Extension of X-11 decomposition with non integer periodicity

`rjd3stl`: STL, MSTL, ISTL, loess 


See next presentation of Anna Smyk


# Other packages


## ggdemetra3
### ggdemetra3 {.allowframebreaks}

Like `ggdemetra` but compatible with `rjdemetra3`: ggplot2 to add seasonal adjustment statistics to your plot,`autoplot()` functions... Also compatible with high-frequency methods (WIP): \footnotesize

```{r ggdemetra3, message = FALSE, warning = FALSE,cache=FALSE, fig.dim = c(6,3)}
library(ggdemetra3)
spec <- spec_x13_default("rsa3") |> 
  set_tradingdays(option = "WorkingDays")
ggplot(data = ipi_c_eu_df, mapping = aes(x = date, y = FR)) +
    geom_line() +
    labs(title = "SA - IPI-FR",
         x = NULL, y = NULL) +
  geom_sa(component = "y_f(12)", linetype = 2,
          spec = spec) + 
  geom_sa(component = "sa", color = "red") +
  geom_sa(component = "sa_f", color = "red", linetype = 2) +
  geom_outlier(geom = "label_repel",
               coefficients = TRUE,
               ylim = c(NA, 65), force = 10,
               arrow = arrow(length = unit(0.03, "npc"),
                             type = "closed", ends = "last"),
               digits = 2)
```

### Plot from an existing model

\footnotesize

```{r ggdemetra3-mod, fig.dim = c(6,3), cache = FALSE}
mod <- rjd3x13::x13(y, spec)
# siratioplot(mod) # SI Ratio plot
autoplot(mod) # autoplot
```

## rjd3filters

### `rjd3filters`

- \bclampe easily create/combine/apply moving averages `moving_average()` (much more general than `stats::filter()`) and study their properties: plot coefficients (`plot_coef()`), gain (`plot_gain()`), phase-shift (`plot_phase()`) and different statics (`diagnostic_matrix()`)

Goal: manipulate moving averages
$$
M_\theta(X_t)=\sum_{k=-p}^{+f}\theta_kX_{t+k} = 
\left(\sum_{k=-p}^{+f}\theta_k B^{-k}\right)X_{t}\text{ with }B^k=X_{t-k}
$$
(Currently in `r fa_r_project` you can only limited forms of MA)


. . .

- \bclampe Lots of complicated and interesting things around trend-cycle estimates


See my presentation at **SAPW 2022** Concurrent Session 5b - Trends
https://community.amstat.org/governmentstatisticssection/conferences/pastconference210/seasonal-adjustment-practitioners-workshop-2022


## rjd3sts and rjd3bench

### rjd3sts and rjd3bench

`rjd3sts` Interface to structural time series and state space models

Several examples available here https://github.com/palatej/test_rjd3sts

`rjd3bench` Benchmarking and temporal disaggregation

Several examples here: https://github.com/palatej/test_rjd3bench



# Conclusion

### Conclusion

With JDemetra+ 3.0, lots of new `r fa_r_project` packages are coming:

- On time series analysis and seasonal adjustment (much faster than standard packages)

- New developments on seasonal adjustment will be available (e.g. high-frequency data)

- Allow to create new trainings thanks to a deeper acces to all the functionalities of JDemetra+


Many ways to contribute:

- Testing it and reporting issues

- Developping new tools (other packages, new functions, etc.)


### Thank you for your attention {.noframenumbering}

Packages \faIcon{r-project}{}:

::::{.columns}
:::{.column width=40%}

\href{https://github.com/palatej/rjd3toolkit}{\faGithub{} palatej/rjd3toolkit}

\href{https://github.com/palatej/rjd3modelling}{\faGithub{} palatej/rjd3modelling}

\href{https://github.com/palatej/rjd3sa}{\faGithub{} palatej/rjd3sa}

\href{https://github.com/palatej/rjd3arima}{\faGithub{} palatej/rjd3arima}

\href{https://github.com/palatej/rjd3x13}{\faGithub{} palatej/rjd3x13}

\href{https://github.com/palatej/rjd3tramoseats}{\faGithub{} palatej/rjd3tramoseats}

\href{https://github.com/palatej/rjdemetra3}{\faGithub{} palatej/rjdemetra3}
:::
:::{.column width=50%}

\href{https://github.com/palatej/rjdfilters}{\faGithub{} palatej/rjdfilters}

\href{https://github.com/palatej/rjd3sts}{\faGithub{} palatej/rjd3sts}

\href{https://github.com/palatej/rjd3stl}{\faGithub{} palatej/rjd3stl}

\href{https://github.com/palatej/rjd3highfreq}{\faGithub{} palatej/rjd3highfreq}

\href{https://github.com/palatej/rjd3bench}{\faGithub{} palatej/rjd3bench}

\href{https://github.com/AQLT/ggdemetra3}{\faGithub{} AQLT/ggdemetra3}
:::
::::

